A comparison of design-based and model-based approaches for finite population spatial data.

### 2022-04-13 inst/manuscript/manuscript.Rmd
review edits through 2.1

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data. In the simulated data, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simulated and real data, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the Generalized Random Tessellation Stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the Simple Random Sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based approaches tend to outperform design-based approaches, even for skewed data where the model-based distributional assumptions are violated. The performance gap between these approaches is small GRTS samples are used but large when SRS samples are used. This suggests that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

Design-based inference; Finite Population Block Kriging (FPBK); Generalized Random Tessellation Stratified (GRTS) algorithm; Local neighborhood variance estimator; Model-based inference; Restricted Maximum Likelihood (REML) estimation; Spatially balanced sampling; Spatial covariance

@@

When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units -- this subset is called a sample. There are two general approaches for using samples to make frequentist statistical inferences about a population: design-based and model-based.  In the design-based approach, inference relies on randomly assigning some population units to be in the sample (random sampling). Alternatively, in the model-based approach, inference relies on distributional assumptions about the underlying stochastic process that generated the sample. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation]. In this manuscript, we compare the design-based and model-based approaches for finite population spatial data.

@@

Spatial data are data that incorporate the locations of the population units into either the sampling or estimation process. @de1990model and @brus1993design give early comparisons of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design-based and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and for Bayesian approaches, see @chan2020bayesian and @hofman2021many).

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
Certainly comparisons between design-based and model-based approaches have been studied in spatial contexts. Our contribution is comparing design-based approaches that incorporate spatial locations into sampling and analysis to model-based approaches. Though the broad comparisons we draw between design-based and model-based approaches generalize to finite and infinite populations, we focus on finite populations.  A finite population contains a finite number of population units (we assume the finite number is known); an example is lakes (treated as a whole with the lake centroid representing location) in the contiguous United States.  An infinite population contains an infinite number of population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for the design-based and model-based approaches to finite population spatial data. In Section \ref{sec:mm}, we describe how we compare performance of the approaches with a simulation study and an analysis of real data that contains mercury concentration in lakes located in the contiguous United States. In Section \ref{sec:results}, we present results from the simulation study and the mercury concentration analysis. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of population units according to a sampling design. A sampling
design assigns a probability of selection to each sample (a subset of population
units). Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. The inclusion probability of a population unit follows by summing each population unit's probability of selection in each sample containing that population unit. Inclusion probabilities are later used to estimate population parameters.

When samples are chosen in a manner such that the layout of sampled units reflects the layout of the population units, we call the resulting sample "spatially balanced." By "reflecting the layout of the population units", we mean that if population units are concentrated in specific areas, the units in the sample should be concentrated in the same areas. Because spatially balanced samples reflect the layout of the population units, they are not necessarily "spread out" in space in some equidistant manner. One approach to selecting spatially balanced samples is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}.

@@

The model-based approach assumes the population is a random realization of a data-generating stochastic process called a superpopulation. Randomness is formally incorporated through distributional assumptions on the superpopulation. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches can implement some form of random sampling, though it is common for model-based approaches to not implement random sampling. When model-based approaches do implement random sampling, the inclusion probabilities are ignored when analyzing the sample (in contrast to the design-based approach, which relies on these inclusion probabilities to analyze the sample).

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable. For example, suppose the realized mean of all population units (the realized population mean) is the variable of interest. Instead of a fixed, unknown mean, we are predicting the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate response values from the same process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures).

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/dvm_comp.jpeg}
  \caption{A visual comparison of the design-based and model-based approaches. In the top row, the design-based approach is highlighted. There is one fixed population with nine population units and three random samples of size four (points circled are those sampled). The response values at each site are fixed, but we obtain different estimates for the mean response in each random sample. In the bottom row, the model-based approach is highlighted. There are three realizations of the same data-generating stochastic process that are all sampled at the same four locations. The data-generating stochastic process has a single mean, but the mean of the nine population units is different in each of the three realizations.}
  \label{fig:fig1}
\end{figure}

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata @walvoort2010r, the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the Generalized Random Tessellation Stratified (GRTS) algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units (replacement units are population units that can be sampled when a population unit originally selected can no longer be sampled).

The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. First the bounding box of the domain is split up into four distinct, equally sized squares called level-one cells. Each level-one is randomly assigned an level-one address of 0, 1, 2, or 3. The set of level-one cells is denoted by $\mathcal{A}_1$ and defined as $\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}$. Within each level-one cell, the inclusion probability for each population unit is summed, and if any of these sums exceed one, a second level of cells is added. Then each level-one cell is split into four distinct, equally sized squares called level-two cells. Each level-two cell is randomly assigned a level-two address of 0, 1, 2, or 3. The set of level-two cells is denoted by $\mathcal{A}_2$ and defined as $\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}$. The inclusion probabilities within each level-two cell are summed, and if any of these sums exceed one, a third level of cells is added. This process continues for $k$ steps, until all level-$k$ cells have inclusion probability sums no larger than one. Then $\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}$.

After determining $\mathcal{A}_k$, it is placed into hierarchical order. Hierarchical order is a numeric order that first sorts $\mathcal{A}_k$ by the level-one addresses from smallest to largest, then sorts $\mathcal{A}_k$ by the level-two addresses from smallest to largest, and so on. For example, $\mathcal{A}_2$ in hierarchical order is the set $\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}$. Because hierarchical ordering sorts by level-one cells, then level-two cells, and so on, population units that have similar hierarchical addresses tend to be nearby one another in space.Next each population unit is mapped to a one-dimensional line in hierarchical order where each population unit's inclusion probability equals its line-length. If a level-$k$ cell has multiple population units in it, they are randomly placed within the cell's respective line segment. A uniform random variable is then simulated in $[0, 1]$ and a systematic sample is selected on the line, yielding $n$ sample points for a sample size $n$. Each element in this systematic sample falls on some population unit's line segment, and thus that population unit is selected in the sample. For further details regarding the GRTS algorithm, see @stevens2004spatially.

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimator for $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th population unit in the sample, $\pi_i$ is the inclusion probability of the $i$th population unit in the sample, and $n$ is the sample size. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

It is also important to quantify the uncertainty in $\hat{\tau}_{ht}$.  @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but these estimators have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically for GRTS samples. Second, these estimators tend to ignore the spatial locations of the population units. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and estimates the variance of $\hat{\tau}$ conditional on the random properties of the GRTS sample -- the idea being that this conditioning should yield a more precise estimate of $\hat{\tau}$. They show that the each observation's contribution to the overall variance is dominated by local variation. Thus the local neighborhood variance estimator is a weighted sum of variance estimates from each observation's local neighborhood. These local neighborhoods contain observation itself and its three nearest neighbors. For more details, see @stevens2003variance.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FPBK next -- for technical details, see @verhoef2008spatial. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal $1 / N$). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u}, let

@@

@@

REWRITE AS SIMPLE RANDOM SAMPLING AND WITHOUT REPLACEMENT

USE SRS / GRTS - DB / MB

ADD LOHR REFERENCE AND FORM FOR SRS VARIANCE WITHOUT REPLACEMENT WITH FPC

We used a simulation study to investigate performance of four sampling-analysis combinations. The first sampling-analysis combination was IRS-Design. In IRS-Design, samples were selected with the Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial locations of the population units, thus the IRS samples were not spatially balanced. In IRS-Design, samples were analyzed using the design-based approach via the Horvitz-Thompson mean estimator and an IRS variance estimator that ignored the spatial locations of the units in the sample. The second sampling-analysis combination was IRS-Model, where samples were selected with the IRS algorithm and analyzed using the model-based approach while estimating the covariance parameters ($\bm{\delta}$) and fixed effects ($\bm{\beta}$ using restricted maximum likelihood (REML). The third sampling-analysis combination was GRTS-Design, where samples were selected with the GRTS algorithm and analyzed using the design-based approach via the Horvitz-Thompson mean estimator and the local neighborhood variance estimator (which does incorporate the spatial locations of the units in the sample). The fourth and final sampling-analysis combination was GRTS-Model, where samples were selected with the GRTS algorithm and analyzed using the model-based approach while estimating the covariance parameters ($\bm{\delta}$) and fixed effects ($\bm{\beta}$ using restricted maximum likelihood (REML). These sampling-analysis combinations are also provided in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS and GRTS samples, equal inclusion probabilities were assumed for all population units. When IRS assumes equal inclusion probabilities for all population units, the algorithm is equivalent to simple random sampling (SRS).

```{r, echo = FALSE, results = "asis"}
tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
       nrow = 2, byrow = TRUE) %>%
  as.table()
row.names(tab) <- c("IRS", "GRTS")
xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Sampling-analysis combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.")
align(xtab1) <- "r|ll"
names(xtab1) <- c("Design", "Model")
print(xtab1, comment = FALSE)
## I really want "Analysis" to appear as a row header and "Sampling Design" 
## to appear as a column header, but this isn't easy to figure out.
## I couldn't figure this out with xtable either so it might need to be
## made "by hand."
```

@@

The mean bias was nearly zero for all four sampling-analysis combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for mean bias in all 36 simulation scenarios are provided in the supporting information.

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data. In the simulated and real data, a variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based approaches tend to outperform design-based approaches, even for skewed data where the model-based distributional assumptions are violated. The performance gap between these approaches is small GRTS samples are used but large when SRS samples are used. This suggests that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

Design-based inference; Finite population block kriging (FPBK); Generalized random tessellation stratified (GRTS) algorithm; Local neighborhood variance estimator; Model-based inference; Restricted maximum likelihood (REML) estimation; Spatially balanced sampling; Spatial covariance

@@

When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units -- this subset is called a sample. There are two general approaches for using samples to make frequentist statistical inferences about a population: design-based and model-based.  In the design-based approach, inference relies on randomly assigning some population units to be in the sample (random sampling). Alternatively, in the model-based approach, inference relies on distributional assumptions about the underlying stochastic process generating the sample. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation, @brus1997random]. In this manuscript, we compare the design-based and model-based approaches for finite population spatial data.

@@

Spatial data are data that have some sort of spatial index, usually via coordinates. @de1990model and @brus1993design give early comparisons of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design-based and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and for Bayesian approaches, see @chan2020bayesian and @hofman2021many).

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
While comparisons between design-based and model-based approaches have been studied in spatial contexts, our contribution is comparing design-based approaches specifically built for spatial data to model-based approaches. Though the broad comparisons we draw between design-based and model-based approaches generalize to finite and infinite populations, we focus on finite populations.  A finite population contains a finite number of population units (we assume the finite number is known); an example is lakes (treated as a whole with the lake centroid representing location) in the conterminous United States.  An infinite population contains an infinite number of population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for design-based and model-based approaches to finite population spatial data. In Section \ref{sec:mm}, we describe how we intend to compare performance of the approaches using simulated and real data. In Section \ref{sec:results}, we present analysis reslts for the simulated and real data. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of population units according to a sampling design. A sampling
design assigns a probability of selection to each sample (subset of population
units). Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. The inclusion probability of a population unit follows by summing each sample's probability of selection over all samples that contain the population unit. Inclusion probabilities are later used to estimate population parameters.

When samples are chosen in a manner such that the layout of sampled units reflects the layout of the population units, we call the resulting sample spatially balanced. By "reflecting the layout of the population units", we mean that if population units are concentrated in specific areas, the units in the sample should be concentrated in the same areas. Because spatially balanced samples reflect the layout of the population units, they are not necessarily spread out in space in some equidistant manner. One approach to selecting spatially balanced samples is the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}.

@@

The model-based approach assumes the population is a random realization of a data-generating stochastic process. Randomness is formally incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches can implement some form of random sampling, though it is common for model-based approaches to sample non-randomly. When model-based approaches do implement random sampling, the inclusion probabilities are ignored when analyzing the sample (in contrast to the design-based approach, which relies on these inclusion probabilities to analyze the sample).

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable. For example, suppose the realized mean of all population units (the realized population mean) is the variable of interest. Instead of a fixed, unknown mean, we are predicting the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate realizations from the same process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures). This figure contrasts the design-based approach with a fixed population and random sampling to the model-based approach with random populations and non-random sampling.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/dvm_comp.jpeg}
  \caption{A visual comparison of the design-based and model-based approaches. In the top row, the design-based approach is highlighted. There is one fixed population with nine population units and three random samples of size four (points circled are those sampled). The response values at each site are fixed. In the bottom row, the model-based approach is highlighted. There are three realizations of the same data-generating stochastic process that are all sampled at the same four locations. The response values at each site are random.}
  \label{fig:fig1}
\end{figure}

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (i.e., Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata @walvoort2010r, the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or on private land that the researcher does not have permission to access.

The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. First the bounding box of the domain is split up into four distinct, equally sized squares called level-one cells. Each level-one cell is randomly assigned a level-one address of 0, 1, 2, or 3. The set of level-one cells is denoted by $\mathcal{A}_1$ and defined as $\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}$. Within each level-one cell, the inclusion probability for each population unit is summed, and if any of these sums exceed one, a second level of cells is added. Then each level-one cell is split into four distinct, equally sized squares called level-two cells. Each level-two cell is randomly assigned a level-two address of 0, 1, 2, or 3. The set of level-two cells is denoted by $\mathcal{A}_2$ and defined as $\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}$. The inclusion probabilities within each level-two cell are summed, and if any of these sums exceed one, a third level of cells is added. This process continues for $k$ steps, until all level-$k$ cells have inclusion probability sums no larger than one. Then $\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}$.

After determining $\mathcal{A}_k$, it is placed into hierarchical order. Hierarchical order is a numeric order that first sorts $\mathcal{A}_k$ by the level-one addresses from smallest to largest, then sorts $\mathcal{A}_k$ by the level-two addresses from smallest to largest, and so on. For example, $\mathcal{A}_2$ in hierarchical order is the set \linebreak $\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}$. Because hierarchical ordering sorts by level-one cells, then level-two cells, and so on, population units that have similar hierarchical addresses tend to be nearby one another in space. Next each population unit is mapped to a one-dimensional line in hierarchical order where each population unit's inclusion probability equals its line-length. If a level-$k$ cell has multiple population units in it, they are randomly placed within the cell's respective line segment. A uniform random variable is then simulated in $[0, 1]$ and a systematic sample is selected on the line, yielding $n$ sample points for a sample size $n$. Each of these sample points falls on some population unit's line segment, and thus that population unit is selected in the sample. For further details regarding the GRTS algorithm, see @stevens2004spatially.

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimator for $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n z_i \pi_i^{-1},
\end{align}
where $z_i$ is the value of the $i$th population unit in the sample, $\pi_i$ is the inclusion probability of the $i$th population unit in the sample, and $n$ is the sample size. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

It is also important to quantify the uncertainty in $\hat{\tau}_{ht}$.  @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but these estimators have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically for GRTS samples. Second, these estimators tend to ignore the spatial locations of the population units. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and estimates the variance of $\hat{\tau}$ conditional on the random properties of the GRTS sample -- the idea being that this conditioning should yield a more precise estimate of $\hat{\tau}$. They show that the contribution from each sample unit (population unit in the sample) to the overall variance is dominated by local variation. Thus the local neighborhood variance estimator is a weighted sum of variance estimates from each sample unit's local neighborhood. These local neighborhoods contain the sample unit itself and its three nearest neighbors among all other sample units. For more details, see @stevens2003variance.

@@

Finite population block kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FPBK next -- for more details, see @verhoef2008spatial. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal $1 / N$). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u}, let

@@

In this section we describe how we used simulated and real data to investigate performance between simple random sampling without replacement (SRS) and GRTS sampling as well as performance between design-based (DB) and model-based (MB) inference. In SRS and GRTS sampling, all population units had equal inclusion probabilities. The important distinction between SRS and GRTS is that SRS ignores spatial locations while sampling but GRTS explicitly incorporates them. Together, the two sampling plans (SRS and GRTS) combined with the two inference approaches (DB and MB) yielded four sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB. For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to estimate means and the commonly-used SRS variance formula [@sarndal2003model; @lohr2009sampling] was used to estimate the variance. This variance formula is given by
\begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation}
where $z_i$ is the $i$th response value, $\bar{z}$ is the mean of all $z_i$, $n$ is the sample size, $N$ is the population size, and $f = (1 - n / N)$ ($f$ is often called the finite population correction factor). For GRTS-DB, the Horvitz-Thompson esetimator was used to estimate means and the local neighborhood variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate means and variances and parameters were estimated using restricted maximum likelihood.

We used simulated data to compare the sampling-inference combinations across many realized populations from the same data-generating stochastic process. With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. We used real data from the 2012 National Lakes Assessment [@USEPA2012NLA] to compare the sampling-inference combinations within a single realized population (which is typically the case in reality). With the real data, we were in control of only the random sampling process.

@@

```{r, echo = FALSE, results = "asis"}
# tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
#        nrow = 2, byrow = TRUE) %>%
#   as.table()
# row.names(tab) <- c("IRS", "GRTS")
# xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Sampling-inference combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.")
# align(xtab1) <- "r|ll"
# names(xtab1) <- c("Design", "Model")
# print(xtab1, comment = FALSE)
# ## I really want "Analysis" to appear as a row header and "Sampling Design" 
# ## to appear as a column header, but this isn't easy to figure out.
# ## I couldn't figure this out with xtable either so it might need to be
# ## made "by hand."
```

@@

The mean bias was nearly zero for all four sampling-inference combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for mean bias in all 36 simulation scenarios are provided in the supporting information.
